---
title: "2.3-Testing_some_stereotypes"
format: html
editor: visual
---

# 2.3 Testing some stereotypes 

## Load the data 

```{r Load the data}
# Load the snoring dataset from the Data folder.
snore <- read.table("Data/snore.txt", header = TRUE)

# Check the structure of the dataset.
str(snore)

# Display the first observations.
head(snore)
```

## Check the categorical variables

```{r Check categorical variables}
# Number of participants
nrow(snore)

# Number of participants according to snoring status
table(snore$snore)

# Number of participants according to sex
table(snore$sex)

# Number of participants according to smoking status
table(snore$tabac)
```

## Are snorers fatter?

To compare body size between snorers and non-snorers, we calculate the Body Mass Index (BMI) from body weight and height, which accounts for differences in body size.

```{r Calculate BMI}

# Calculate height in metres
snore$height_m <- snore$height / 100

# Calculate Body Mass Index (BMI)
snore$BMI <- snore$weight / (snore$height_m^2)

# Check the calculated BMI values
head(snore[, c("weight", "height", "BMI")])
```

We calculate the mean and median BMI in each snoring group to describe and compare the central tendency of the two groups.

```{r BMI Means}

# Mean BMI for each snoring group
tapply(snore$BMI, snore$snore, mean)

# Median BMI for each snoring group
tapply(snore$BMI, snore$snore, median)
```

We visualize the BMI distributions in snorers and non-snorers to assess their variability and identify possible differences between the two groups.

```{r Compare BMI}

# Compare BMI distributions between snorers and non-snorers
ggplot(snore, aes(x = snore, y = BMI)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1) +
  labs(
    x = "Snoring status",
    y = "Body Mass Index (BMI)"
  )
```

We use a Mann–Whitney U test to determine whether BMI differs significantly between snorers and non-snorers.

```{r Mann-Whitney U test BMI}
# Perform a Mann–Whitney U test
# H0: BMI does not differ between snorers and non-snorers
# H1: BMI differs between snorers and non-snorers

wilcox.test(BMI ~ snore, data = snore)
```

The p-value is compared with the significance level of 0.05 to determine whether the observed difference in BMI between the two groups is statistically significant.

#### Results : 

The mean BMI was 27.27 kg/m² in non-snorers and 27.11 kg/m² in snorers, while the median BMI was 28.07 and 28.08 kg/m², respectively. The Mann–Whitney U test showed no statistically significant difference in BMI between the two groups (W = 1128.5, p = 0.951). Therefore, these data do not provide evidence that snorers have a different BMI from non-snorers.

## Do snorers drink more ?

We first calculate the mean and median alcohol consumption in each snoring group to describe and compare their alcohol consumption.

```{r Alcohol consumption Means}
# Mean alcohol consumption for each snoring group
tapply(snore$alcool, snore$snore, mean)

# Median alcohol consumption for each snoring group
tapply(snore$alcool, snore$snore, median)
```

We visualize alcohol consumption in snorers and non-snorers to compare the distributions and identify possible differences or extreme values.

```{r Alcohol consumption Boxplot}
# Compare alcohol consumption between snorers and non-snorers
ggplot(snore, aes(x = snore, y = alcool)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1) +
  labs(
    x = "Snoring status",
    y = "Alcohol consumption"
  )
```

We use a Mann–Whitney U test to determine whether alcohol consumption differs significantly between snorers and non-snorers.

```{r Mann–Whitney U test3}

# Perform a Mann–Whitney U test
# H0: alcohol consumption does not differ between snorers and non-snorers
# H1: alcohol consumption differs between snorers and non-snorers

wilcox.test(alcool ~ snore, data = snore)
```

### Results : 

Alcohol consumption was higher among snorers (mean = 4.03, median = 4) than among non-snorers (mean = 2.37, median = 0). For the statistical test part, since p = 0.008594 \< 0.05, we reject the null-hypothesis H0. A statistically significant difference in alcohol consumption was observed between snorers and non-snorers (Mann–Whitney U test, W = 788, p = 0.0086).
