---
title: "_02-prepost"
format: html
---

## 2.2 Effect of treatment over time

### 2.2.1 Looking at the data

Loading the necessary libraries

```{r}
# Load standard libraries
library(tidyverse)
```

Loading the data

```{r Loading the data and converting the data to columns}

# Loading the data
data_long <- read_csv("../Data/PrePost.csv")

# Adding Subject IDs (assuming 8 patients in the exact same order)
# and convert 'time' to a factor so 'Before' always plots first
data_long <- data_long %>%
  mutate(ID = rep(1:8, times = 2)) %>%
  mutate(time = factor(time, levels = c("Before", "After")))

# Pivot to wide format to easily calculate the difference
data_wide <- data_long %>%
  pivot_wider(names_from = time, values_from = perf) %>%
  mutate(difference = After - Before)

head(data_wide)
```

Variable Descriptions

| Variable | Description | Unit | Possible values | Expected change |
|:--------------|:--------------|:--------------|:--------------|:--------------|
| `perf` | Physical performance score of the patient | not given in the file | positive values; here 125 to 148 | ↑ after rehabilitation |
| `time` | Moment of the measurement | — | `Before`, `After` | — (factor) |

### **2.2.2 The analysis**

#### Assessing normality

First will check the normality by Shapiro-Wilk Normality Test, before deciding using Parametric or non-parametric test.

```{r checking normality}
# Check normality assumption
shapiro_test <- shapiro.test(data_wide$difference)
print(shapiro_test)
```

A Shapiro-Wilk test was conducted on the differences between post- and pre-treatment performance scores to evaluate the assumption of normality. The results show that the distribution of score differences does not significantly deviate from a normal distribution (W=0.899,p=0.283). Because p\>0.05, the assumption of normality is met, validating the choice of a parametric paired t-test for the main analysis.

#### 
